Improved Direct-Control Reinforcement Learning for DDoS Prevention

### 2019-02-06 glass-talk/glass-talk.tex
Finished presentation

\begin{frame}{Reinforcement Learning: How it Works (II)}
% make math easy
\begin{subequations}
	We can keep updating the policy by updating $\wvec{}$ as we receive state ($S_t, S_{t+1}$), action ($A_t, A_{t+1}$), reward information ($R_{t+1}$).
	The \alert{semi-gradient Sarsa algorithm} works here, starting from $\wvec{0}=\bm{0}$:
	\begin{gather}
	\delta_t = R_{t+1} + \gamma \acval{S_{t+1}}{A_{t+1}}{\wvec{t}} - \acval{S_t}{A_t}{\wvec{t}},\\
	\bm{w}_{t+1} = \bm{w}_{t} + \alpha \delta_t \nabla_{\wvec{}}{\acval{S_t}{A_t}{\wvec{t}}}
	\end{gather}
	Our \alert{policy} is then to choose whichever action has the highest estimated value in the current state, or choose at random with probability $\epsilon$ ($\epsilon$-greedy policy).
\end{subequations}
\end{frame}

@@

\begin{frame}{Multiagent RL for DDoS prevention}
%	Network Model \footcite{DBLP:journals/eaai/MalialisK15}
%	
%	Pushback \footcite{DBLP:journals/ccr/MahajanBFIPS02a}
%	
%	Group reward functions etc.
%
%	Its weaknesses? Strong assumptions about what knowledge the learners really have...

	\begin{itemize}
		\item Yet existing RL-based DDoS defences like MARL\footcite{DBLP:journals/eaai/MalialisK15} haven't been successful---why?
		% How? Violation of ISP-like criterion,

@@

		\item \alert{Actions not granular}---many hosts punished accidentally.
		\item \alert{Action affects congestion-aware traffic disproportionately}---TCP, QUIC \emph{actively slow down} if packets lost.

@@

\begin{frame}{Multiagent RL (MARL) for DDoS prevention}
\begin{columns}
	\begin{column}{0.45\linewidth}
		\begin{itemize}
			\item Algorithm: Semi-gradient Sarsa, linear fn approx, $\epsilon$-greedy selection.
			\item Actions: Drop [0, 10, ... 90]\si{\percent} upstream traffic.
			\item State: load vectors of agent and parents ($\mathbb{R}^4$) $\rightarrow$ tile-coded (fixed-weight binary vector).
			\item Rewards: \num{-1} if $\mathit{load} > U_s$, else fraction of surviving legit traffic.
		\end{itemize}
	\end{column}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\centering
			\resizebox{\linewidth}{!}{
				\begin{tikzpicture}[
				texts/.style = {text=black},
				labeltexts/.style = {text=uofgsandstone},
				treeline/.style = {draw=uofgburgundy},
				treenode/.style = {texts, circle, centered, fill=white, treeline},
				load/.style = {fill=uofgcobalt},
				external/.style = {fill=uofgrust},
				hideline/.style = {draw=uofgsandstone!40!white},
				hidenode/.style = {treenode, hideline},
				grow'=right
				]
				\node[treenode, label={[texts]above:Server}] (root) {}
				child [treeline] { node [treenode, load, label={[texts]above:Core}] (sswitch) {}
					child [treeline] { node [treenode, load, label={[texts]above:Leader}] (teaml) {} 
						child [treeline] { node [treenode, load, label={[texts]above:Intermediate}] (inter) {}
							child [treeline] { node [treenode, load, label={[texts]above:Agent/Egress}] (agent) {}
								child [treeline] { node [treenode, external] (extern) {}
									child [treeline] { node [treenode, external, label={[texts]above:Host}] (host) {} }
									child [hideline] { node [hidenode] (endhost) {} }
								}
							}
							child [hideline] { node [hidenode] (endagent) {} }
						}
						child [hideline] { node [hidenode] (endinter) {} }
					}
					child [hideline] { node [hidenode] (endteaml) {} }
					edge from parent
					node[below, labeltexts] {$U_s$}
				};

@@

@@

\begin{frame}{Adapting MARL}
\begin{itemize}
	\item Prior work with these techniques has two key flaws:
	\begin{itemize}
		\item Each action can (and will!) map to many flows (particularly in real networks), increasing the \alert{likelihood} of collateral damage.
		\item The majority of internet traffic is congestion-aware, increasing the \alert{cost} of collateral damage.
	\end{itemize}
	\item Naturally, we aim then to take decisions \emph{per-flow}.
\end{itemize}
\end{frame}

@@

\begin{frame}{Feature Performance---UDP}
\begin{figure}
	\centering
	%	\begin{subfigure}{0.32\linewidth}
	%		\includegraphics[width=\linewidth]{../plots/ftprep-good}
	%	\end{subfigure}
	%	\begin{subfigure}{0.32\linewidth}
	%		\includegraphics[width=\linewidth]{../plots/ftprep-good-2}
	%	\end{subfigure}
	%	\begin{subfigure}{0.32\linewidth}
	%		\includegraphics[width=\linewidth]{../plots/ftprep-good-3}
	%	\end{subfigure}
	\includegraphics[width=0.8\linewidth]{../plots/ftprep-box}
%	\vspace{-1.2cm}
	\caption{
		Learned performance of Marl++ agents when benign traffic is UDP-like, using only a single feature as a basis for decisions. Uncap.
		\label{fig:udp-feature-plots}
	}
\end{figure}
\end{frame}

\begin{frame}{Feature Performance---TCP}
\begin{figure}
	\centering
	%	\begin{subfigure}{0.32\linewidth}
	%		\includegraphics[width=\linewidth]{../plots/ftprep-good}
	%	\end{subfigure}
	%	\begin{subfigure}{0.32\linewidth}
	%		\includegraphics[width=\linewidth]{../plots/ftprep-good-2}
	%	\end{subfigure}
	%	\begin{subfigure}{0.32\linewidth}
	%		\includegraphics[width=\linewidth]{../plots/ftprep-good-3}
	%	\end{subfigure}
	\includegraphics[width=0.8\linewidth]{../plots/ftprep-tcp-cap-box}
%	\vspace{-1.2cm}
	\caption{
		Learned performance of Marl++ agents when benign traffic is TCP-like, using only a single feature as a basis for decisions.
		\label{fig:udp-feature-plots-cap}
	}
\end{figure}
\end{frame}

@@

\section{Evaluation}
\begin{frame}{UDP traffic performance}
\begin{columns}
	\begin{column}{0.7\linewidth}
	\begin{figure}
		\centering
		\includegraphics[width=0.95\linewidth]{../plots/udp-box}

@@

\begin{frame}{TCP traffic performance}
\begin{columns}
	\begin{column}{0.3\linewidth}
		\begin{itemize}
			\item Harder to see statistically significant advantage until large $n$.
			\item SPF has low variance as expected, but performs poorly until large $n$.
		\end{itemize}
	\end{column}
	\begin{column}{0.7\linewidth}
		\begin{figure}
			\centering
			\includegraphics[width=0.95\linewidth]{../plots/tcp-box}

@@

\begin{frame}{TCP traffic (n=16)---training in detail}
	\begin{figure}
		\centering
		\includegraphics[width=0.8\linewidth]{../plots/tcp-16-single}

@@

\begin{frame}{Other environments}
\begin{itemize}
	\item Expansion to other topologies.
	\begin{itemize}
		\item Likely not viable or sensible to map this overlay topology to most networks: many sources/sinks, multiple paths make global state useless.
		\item All we need to change is global state.
	\end{itemize}
	\item So flip the problem on its head...
	\item This will enable direct comparison against other approaches.
\end{itemize}
\end{frame}

\begin{frame}{Other attacks}
\begin{itemize}
	\item More ``old-school'' attacks like TCP SYN/ACK floods, application-level attacks against emulated software...
	\item In AS-like networks/topologies, we can examine \alert{Link Flooding Attacks}.
	\item Crucially, need several of these to evaluate against attackers with \alert{evolving strategies} (and show generalisation beyond amplification attacks).
\end{itemize}
\end{frame}

>>>>

\begin{frame}{Reinforcement Learning: How it Works (II)}
% make math easy
\begin{subequations}
	We can keep updating the policy by updating $\wvec{}$ as we receive state ($S_t, S_{t+1}$), action ($A_t, A_{t+1}$), reward information ($R_{t+1}$).
	The \alert{semi-gradient Sarsa algorithm} works here, starting from $\wvec{0}=\bm{0}$:
	\begin{gather}
	\delta_t = R_{t+1} + \gamma \acval{S_{t+1}}{A_{t+1}}{\wvec{t}} - \acval{S_t}{A_t}{\wvec{t}},\\
	\bm{w}_{t+1} = \bm{w}_{t} + \alpha \delta_t \nabla_{\wvec{}}{\acval{S_t}{A_t}{\wvec{t}}}
	\end{gather}
	given $\alpha, \gamma \in [0,1]$.

	Our \alert{policy} is then to choose whichever action has the highest estimated value in the current state, or choose at random with probability $\epsilon$ ($\epsilon$-greedy policy).
\end{subequations}
\end{frame}

@@

\begin{frame}{Multiagent RL (MARL) for DDoS prevention}
%	Network Model \footcite{DBLP:journals/eaai/MalialisK15}
%	
%	Pushback \footcite{DBLP:journals/ccr/MahajanBFIPS02a}
%	
%	Group reward functions etc.
%
%	Its weaknesses? Strong assumptions about what knowledge the learners really have...

	\begin{itemize}
		\item There are existing RL-based DDoS defences like MARL\footcite{DBLP:journals/eaai/MalialisK15}.
		% How? Violation of ISP-like criterion,

@@

@@

\begin{frame}{Multiagent RL (MARL) for DDoS prevention}
\begin{columns}
	\begin{column}{0.45\linewidth}
		\begin{itemize}
			\item Algorithm: Semi-gradient Sarsa, linear fn approx, $\epsilon$-greedy selection.
			\item Actions: Drop [0, 10, ... 90]\si{\percent} upstream traffic.
			\item State: load vectors of agent and parents ($\mathbb{R}^4$) $\rightarrow$ tile-coded.
			\item Rewards: \num{-1} if $\mathit{load} > U_s$, else fraction of surviving legit traffic.
		\end{itemize}
	\end{column}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\centering
			\resizebox{\linewidth}{!}{
				\begin{tikzpicture}[
				texts/.style = {text=black},
				labeltexts/.style = {text=uofgsandstone},
				treeline/.style = {draw=uofgburgundy},
				treenode/.style = {texts, circle, centered, fill=white, treeline},
				load/.style = {fill=uofgcobalt},
				external/.style = {fill=uofgrust},
				hideline/.style = {draw=uofgsandstone!40!white},
				hidenode/.style = {treenode, hideline},
				grow'=right
				]
				\node[treenode, label={[texts]above:Server}] (root) {}
				child [treeline] { node [treenode, load, label={[texts]above:Core}] (sswitch) {}
					child [treeline] { node [treenode, load, label={[texts]above:Leader}] (teaml) {} 
						child [treeline] { node [treenode, load, label={[texts]above:Intermediate}] (inter) {}
							child [treeline] { node [treenode, load, label={[texts]above:Agent/Egress}] (agent) {}
								child [treeline] { node [treenode, external] (extern) {}
									child [treeline] { node [treenode, external, label={[texts]above:Host}] (host) {} }
									child [hideline] { node [hidenode] (endhost) {} }
								}
							}
							child [hideline] { node [hidenode] (endagent) {} }
						}
						child [hideline] { node [hidenode] (endinter) {} }
					}
					child [hideline] { node [hidenode] (endteaml) {} }
					edge from parent
					node[below, labeltexts] {$U_s$}
				};

@@

\begin{frame}{Multiagent RL (MARL) for DDoS prevention}
%	Network Model \footcite{DBLP:journals/eaai/MalialisK15}
%	
%	Pushback \footcite{DBLP:journals/ccr/MahajanBFIPS02a}
%	
%	Group reward functions etc.
%
%	Its weaknesses? Strong assumptions about what knowledge the learners really have...

\begin{itemize}
	\item Yet existing RL-based DDoS defences like MARL haven't been successful---why?

	\item \alert{Actions not granular}---many hosts punished accidentally.
	\item \alert{Action affects congestion-aware traffic disproportionately}---TCP, QUIC \emph{actively slow down} if packets lost.
\end{itemize}
\end{frame}

@@

\begin{frame}{Adapting MARL}
\begin{itemize}
	\item Prior work with these techniques has two key flaws:
	\begin{itemize}
		\item Each action can (and will!) map to many flows (particularly in real networks), increasing the \alert{likelihood} of collateral damage.
		\item The majority of internet traffic is congestion-aware, increasing the \alert{cost} of collateral damage.
	\end{itemize}
	\item Naturally, we aim then to take decisions \emph{per-flow} (or per-source, to be precise).
\end{itemize}
\end{frame}

@@

\begin{frame}{Feature Performance---UDP}
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{../plots/ftprep-box-pres}
%	\vspace{-1.2cm}
	\caption{
		Learned performance of Marl++ agents when benign traffic is UDP-like, using only a single feature as a basis for decisions. Uncap.
		\label{fig:udp-feature-plots}
	}
\end{figure}
\end{frame}

\begin{frame}{Feature Performance---TCP}
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{../plots/ftprep-tcp-cap-box-pres}
%	\vspace{-1.2cm}
	\caption{
		Learned performance of Marl++ agents when benign traffic is TCP-like, using only a single feature as a basis for decisions.
		\label{fig:udp-feature-plots-cap}
	}
\end{figure}
\end{frame}

@@

\section{Evaluation}
\begin{frame}{UDP traffic performance}
\begin{columns}
	\begin{column}{0.7\linewidth}
	\begin{figure}
		\centering
		\includegraphics[width=0.95\linewidth]{../plots/udp-box-pres}

@@

\begin{frame}{TCP traffic performance}
\begin{columns}
	\begin{column}{0.3\linewidth}
		\begin{itemize}
			\item Harder to see statistically significant advantage until large $n$.
			\item All have very high variance (due to poor actions on congestion-aware protocols), but SPF narrows it down.
		\end{itemize}
	\end{column}
	\begin{column}{0.7\linewidth}
		\begin{figure}
			\centering
			\includegraphics[width=0.95\linewidth]{../plots/tcp-box-pres}

@@

\begin{frame}{TCP traffic (n=16)---training in detail}
	\begin{figure}
		\centering
		\includegraphics[width=0.8\linewidth]{../plots/tcp-16-single-pres}

@@

\begin{frame}{Other environments}
\begin{itemize}
	\item Expansion to other topologies.
	\begin{itemize}
		\item Likely not viable or sensible to map this overlay topology to most networks: many sources/sinks, multiple paths make global state useless.
		\item \alert{Then what we need to change is global state!}
	\end{itemize}
	\item So flip the problem on its head: even in multipath networks \emph{we know what path a flow will follow} (i.e., via ECMP).
	\begin{itemize}
		\item So, draw global state from the \emph{path}: endpoints + tertiles (still in $\mathbb{R}^{4}$).
		\item Cost: possibly more network measurements, concept of teams no longer valid (so lose info from reward functions).
	\end{itemize}
	\item This will enable direct comparison against other approaches.
\end{itemize}
\end{frame}

\begin{frame}{Other attacks}
\begin{itemize}
	\item More ``old-school'' attacks like TCP SYN/ACK floods, application-level attacks against emulated software...
	\item In AS-like networks/topologies, we can examine \alert{Link Flooding Attacks}.
	\item Crucially, need several of these to evaluate against attackers with \alert{evolving strategies} (and show generalisation beyond amplification attacks).
	\begin{itemize}
		\item Evolving attackers are \emph{the norm}\footcite{DBLP:conf/spw/KangGS16}, though few works detail \emph{how} they evolve.
	\end{itemize}
\end{itemize}
\end{frame}
